---
title: "HW_2"
format: html
editor: visual
mainfont: "Arial"
editor: visual
author: "Megan Baker and Rebeca Adam"
date: "2026-05-04"
execute:
  echo: true
  warning: false
  message: false
---

## Libraries setup

```{r}
# Clear the environment
rm(list=ls())

# Install "dataRetreival"
install.packages("dataRetrieval")

# Libraries
library(tidyverse)
library(here)
library(ggplot2)
library(lubridate)
library(dataRetrieval)
library(tidyverse)
```

## Pasted from assignment

```{r}
siteNumbers <- c("14211720", "14211010", "14166000")
parameterCd <- "00010"  # water temperature
startDate <- "2010-01-01"
endDate <- "2023-12-31"

temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)
```

```{r}
# rename columns for clarity
colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")
# extract some date info for use later
temp_data$month = month(temp_data$Date, label=TRUE)
temp_data$year = year(temp_data$Date)

# check by plotting
ggplot(temp_data, aes(x=Date, y=temp_C, color=site_no)) +
  geom_line() +
  labs(title="Stream Temperature at Willamette River Sites",
       x="Date", y="Temperature (°C)", color="Site Number") +
  theme_minimal()
```

```{r}
# filter for July temperatures
jan_data <- subset(temp_data, month == "Jan")
# fit a GLM model to see if there is a trend in July temperatures over time, separated by site
jan_model <- glm(temp_C ~ year*site_no, data = jan_data)
summary(jan_model)
```

```{r}
names(jan_data)[5] <- "A"

jan_summary <- jan_data |>
  group_by(site_no) |>
  summarize(
    mean_temp = mean(temp_C, na.rm = TRUE),
    max_temp = max(temp_C, na.rm = TRUE),
    min_temp = min(temp_C, na.rm = TRUE),
    .groups = "drop"
  )

jan_summary
```

```{r}
names(temp_data)[5] <- "A"

jan_data <- temp_data |>
  filter (month == "Jan")|>
  mutate(site_no = as.factor(site_no))
```

#create csv

```{r}
write_csv(temp_data, "willametter_stream_temp.csv")

```

```{r}
ggplot(jan_data, aes(x = Date, y = temp_C, color = site_no)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  labs(
    title = "January Stream Temp Trends at Willamette River Sites",
    x = "Date",
    y = "Temperature (°C)",
    color = "Site Number"
  ) +
  theme_minimal()
```

Analysis of January stream temperatures across three Willamette River sites from 2010 to 2023 reveals modest but notable warming trends. Site 14166000 shows the warmest mean January temperature (6.54°C), followed closely by site 14211720 (6.52°C), while site 14211010 runs notably cooler on average (5.33°C), though all three sites share a similar maximum range of 7.8–9.1°C. The GLM results indicate a positive temperature trend at the reference site (14166000), though this slope is only marginally significant (p = 0.058). Sites 14211010 and 14211720 show steeper warming slopes with site 14211720's steeper trend reaching statistical significance (p = 0.028). The dashed trend lines in the updated plot make this pattern visible: all three sites trend slightly upward over the period, with the two upstream sites warming at a faster rate than the reference site. The sharp downward spikes visible around 2017 for site 14211010 are likely outliers and may be a caveat. Overall, while the warming signal in January is relatively subtle compared to warmer months, the consistent positive direction across all three sites is ecologically meaningful and warrants continued monitoring.
